Lab 2 Final Report: COVID-19 Case Rate vs Population Demographics and Mask Policy

### 2020-12-03 Jackson_Liu_Temlock_Zhang_Lab2_Draft_Report.Rmd
Updated omitted variable section.

One of the variables that was hypothesized as being potentially influential is a ratio of a state's urban to rural population. Currently, the dataset in use contains population information as averaged across each state as a whole. While population density is potentially useful, it was found to not be significant in the models. This may be due to states which large urban centers, where population density woukd be high, losing their impact due to also containing large areas of land that is sparsely populated. One example of this would be New York, which contains several of the most densely populated counties in the country, but also large rural areas. This variable would also be explanatory of the population density variable as a whole, where a higher urban to rural ratio would result in a higher statewide population density generally. With a greater fraction of urban residents, people would also be less likely to be able to social distance as effectively, increasing the per capita case rate. Therefore, the coefficient for population density in the original model would be larger than a model where this ratio is included, resulting in a bias that is positive and away from zero.

@@

Knowledge of the median income for a state would give an indication of the relative job status for a working person in that state. This would be important for the case rate to understand about how many people work in low paying jobs that are more frequently in close contact with others (such as the service industry, gig industry, etc.). Higher median income may imply that more workers are able to function remotely with greater social distancing. A higher median income would then be negatively related to the covid case rate. It would also be explanatory in a way that makes the poverty percentage variable less impactful, where either of these variables may be considered as proxies for one another. Because inclusion of the median income variable would led to a lower magnitude poverty percentage variable, this would cause a positive omitted variable bias away from zero.

@@

This variable could also be partially explained by the senior percentage variable, because of the gap in gender ratios across generations and generally as people get older. Because men have lower life expectancies, older groups of people living in the US would have higher ratios of women to men. Therefore, introducing the gender ratio variable into the model would lower the value of the coefficient to the senior percentage variable. This would be a positive omitted variable bias away from zero.

@@

A variable capturing the mask use among the population would likely reduce the explanatory power of the state mask mandate variable. This would cause the bias of the omitted variable to be positive and away from zero.

@@

The economic standing of a state not only affects the individual households of its population, but also includes how it interacts with other states and how its peoples' lives might have been economically changed by the pandemic. One way that a state may increase its number of covid cases is both from tourists or others visiting during the pandemic, but also if its own population must work in close contact with others. In addition, if certain key industries to a state, such as tourism, layed off many workers, they would likely have to take lower paying close contact jobs such as in the service or gig economy in order to continue having a livelihood.

This effect can be included with a variable that demonstrates how much of a state's GDP is from tourism. It would be related to other economic indicator variables, such as the percent of the population living in poverty. While poverty is a broader phenomena, the number of people in poverty in the pandemic when the data was taken would be potentially higher due to negative effects in the tourism industry. Therefore, it is expected that states with a greater focus on tourism may also have more people in poverty in the pandemic. This would take some of the explanatory power of the poverty variable away, resulting in a positive omitted variable bias away from zero.

>>>>

One of the variables that was hypothesized as being potentially influential is a ratio of a state's urban to rural population. Currently, the dataset contains population information as averaged across each state as a whole. While population density is potentially useful, it was found to not be significant in the models. This may be due to states which have large urban centers, where population density woukd be high, losing their impact due to also containing large areas of land that is sparsely populated. One example of this would be New York, which contains several of the most densely populated counties in the country, but also large rural areas.

An urban to rural ratio would be explanatory of the population density variable as a whole, where a higher urban to rural ratio would result in a higher statewide population density generally. With a greater fraction of urban residents, people would also be less likely to be able to social distance as effectively, increasing the per capita case rate. Therefore, the coefficient for population density in the original model would be larger and more positive than a model where this ratio is included, resulting in a bias that is positive and away from zero.

@@

Knowledge of the median income for a state would give an indication of the relative job status for a working person in that state. This would be important for the case rate to understand about how many people work in low paying jobs that are more frequently in close contact with others (such as the service industry, gig industry, etc.). Higher median income may imply that more workers are able to function remotely with greater social distancing. A higher median income would then be negatively related to the covid case rate. It would also be explanatory in a way that makes the poverty percentage variable less impactful, where either of these variables may be considered as proxies for one another. Because inclusion of the median income variable would led to a lower magnitude poverty percentage variable, and that the coefficient of the poverty percentage variable is positive, this would cause a positive omitted variable bias away from zero.

@@

This variable could also be partially explained by the senior percentage variable, because of the increase of women in gender ratios generally as people get older. Therefore, introducing the gender ratio variable into the model would lower the value of the coefficient to the senior percentage variable to be less negative. This would be a negative omitted variable bias away towards zero.

@@

A variable capturing the mask use among the population would likely reduce the explanatory power of the state mask mandate variable to be less negative. This would cause the bias of the omitted variable to be negative and towards from zero.

@@

The economic standing of a state not only affects the individual households of its population, but also includes how it interacts with other states and how its peoples' lives might have been economically changed by the pandemic. One way that a state may increase its number of covid cases is both from tourists or others visiting during the pandemic and bringing disease, but also if its own population must work in close contact with others serving them in the tourism industry. In addition, if certain key industries to a state, such as tourism, layed off many workers, they would likely have to take lower paying close contact jobs such as in the service or gig economy in order to continue having a livelihood. This would further lead to a higher covid case rate

These effects may be included with a variable that demonstrates how much of a state's GDP is from tourism. It would be related to other economic indicator variables, such as the percent of the population living in poverty. While poverty is a broader phenomena, the number of people in poverty in the pandemic could be potentially higher due to negative growth of the tourism industry. Therefore, it is expected that states with a greater focus on tourism may also have more people in poverty in the pandemic. This would take some of the explanatory power of the poverty variable away, resulting in its model coefficient becoming less positive. This would result in a positive omitted variable bias away from zero.
